Modelling last glacial cycle ice dynamics in the Alps

### 2017-11-03 articles/cycle/alps-cycle.tex
Apply minor corrections from Guillaume.
I have left some apart to think what to do.

    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences and
    self-sustained ice domes. Finally, the Last
    Glacial Maximum advance, often considered synchronous, is here modelled as
    a time-transgressive event, with some glacier lobes reaching their maximum
    as early as 27\,ka, and some as late as 21\,ka. Modelled ice thickness is
    about 900\,m higher than observed trimline elevations, yet our simulation
    predicts little erosion at high elevation due to cold ice conditions.

@@

    From the mid-twentieth
    century, palaeoclimate records extracted from deep sea sediments and
    ice cores have provided a much more detailed picture of the Earth
    environmental history \citep[e.g.,][]{Emiliani.1955,
    Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
    indicating neither one, nor four, but rather tens of glacial and
    interglacial periods. During the last 800000 years (800\,ka), these
    \emph{glacial cycles}
    have succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
    Augustin.etal.2004}. Nevertheless, this global signal is largely governed
    by the North American and Eurasian ice sheet complexes. It is thus unclear
    whether glacier advances and retreats in the Alps were in pace with global
    sea-level fluctuations.

    Besides, the landform record is typically sparse in time and, most often,
    spatially incomplete. Palaeo-ice sheets did not leave a continuous imprint
    on the lanscape, and much of this evidence has been overprinted by
    subsequent glacier re-advances and other geomorphological processes
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2006, Kleman.etal.2010}. Dating
    uncertainties typically increase with age, such that dated reconstructions
    are strongly biased towards earlier glaciations \citep{Heyman.etal.2011}.
    The European Alps have been studied in more detail than any other glacial
    region, but they are no exception to these rules. Although sparse
    geological traces indicate that the last glacial cycle may have comprised
    two or three cycles of glacier growth and decay \citep{Preusser.2004,
    Ivy-Ochs.etal.2008}, most glacial features currently left on the foreland
    present a record of the last major glaciation of the Alps, dating from the
    \emph{Last Glacial Maximum} (LGM).

@@

    These simulations use an horizontal resolution of 2\,km. The vertical grid
    consists of 31~temperature layers in the bedrock and up to 126~enthalpy
    layers in the ice, corresponding to vertical resolutions of 100 and
    40\,\unit{m}, respectively.

@@

    For the three palaeo-temperature records and the two palaeo-precipitation
    parametrizations used, the model yields significant ice volume build-up
    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}b), corresponding to documented
    glaciation periods in the Alps \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
    All simulations also yield important glaciations during MIS~5 and 3, but
    their timing and amplitude varies significantly depending on the climate
    forcing used. All six simulations overestimate
    ice cover during the Younger Dryas \citep[cf. e.g.,][]{Ivy-Ochs.etal.2009}.
    This is partly because the 2\,km resolution is too coarse to resolve
    Younger Dryas Alpine glaciers, but the large total ice volume
    overestimation resulting from the GRIP temperature forcing
    (Fig.~\ref{fig:timeseries}b, blue curves) can not be explained by
    resolution alone.

    Geological data from the best documented Alpine piedmont lobes indicate
    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
    potentially re-advanced to within close reach of this maximum extent, until
    as late as 22 to 17\,ka \citep[Fig.~5]{Wirsig.etal.2016}. This is in very
    good agreement with the result of the EPICA simulations, which yield an
    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
    reductions) followed by a standstill until 17.3\,ka
    (Fig.~\ref{fig:timeseries}b, blue curves). The GRIP palaeo-temperature
    forcing, on the other hand, yield two distinct total ice volume maxima
    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
    a late LGM peak ice volume at 16.5/15.7\,ka several thousand years after
    the dated geological evidence.

@@

    Beside this general pattern, there exist differences in glaciated area
    depending on the palaeo-climate forcing used. The MD01-2444, and to a
    greater extent, the GRIP, palaeo-temperature records, tend to overestimate
    MIS~2 ice cover on all peripheral ranges that surround the Alps
    (Fig.~\ref{fig:footprints}a and c). This is particularly true when
    precipitation corrections are applied (bright colours). In fact, both
    palaeo-temperature records have a higher temperature variability and
    contain brief spells of cold temperatures, which are too short to develop a
    fully grown Alpine ice sheet, but long enough to build up ice cover on a
    smaller scale on these peripheral ranges (Fig.~\ref{fig:timeseries}). On
    the other hand, peripheral glaciation modelled in the EPICA simulation
    (Fig.~\ref{fig:footprints}b) is in relatively good agreement with the
    geomorphological reconstructions.

    The modelled extent of glaciation during MIS~4 depicts a more pronounced
    sensitivity to the choice of palaeo-climate forcing used. Using the GRIP
    and MD01-2444 palaeo-temperature records, Alpine glaciers are modelled to
    extend well beyond the reach of documented LGM end-moraines and glacial
    erratic terrain (Fig.~\ref{fig:footprints}d and f). Because, in both cases,
    the modelled glaciation extent corresponds to brief cold spells in the
    palaeo-temperature forcing (Fig.~\ref{fig:timeseries}a),
    palaeo-precipitation reductions greatly reduce the excessive modelled ice
    cover (Figs.~\ref{fig:timeseries}b and \ref{fig:footprints}d and f, light
    colours). However, with or without palaeo-precipitation corrections,
    the GRIP and MD01-2444 forcing yield
    modelled ice extents (Fig.~\ref{fig:footprints}d and f) and volumes
    (Fig.~\ref{fig:timeseries}b) higher during MIS~4 than
    during MIS~2. This is in contradiction with geological evidence
    \citep{Preusser.2004, Ivy-Ochs.etal.2008}.

    The EPICA palaeo-temperature forcing, on the other hand, yields a MIS~4
    glaciation that is only slightly less expensive than the MIS~2 glaciation
    (Fig.~\ref{fig:footprints}e). The sensitivity of the glaciated area to
    palaeo-precipitation reductions is mostly limited to the southern terminal
    lobes of central-Alpine glaciers, which are slightly less extensive during
    MIS~2 and 4 (Fig.~\ref{fig:footprints}b and e).

    Based on the above considerations on timing of the LGM during MIS~2, and
    modelled ice
    extent during MIS~4, we chose EPICA as our optimal palaeo-temperature
    record for a more detailed and higher-resolution comparison of modelled
    glacier dynamics to available geological evidence
    (Sect.~\ref{sec:results}). As a conservative approach in regard to the
    rapid ice volume fluctuations, we choose to include palaeo-precipitation
    corrections in the following comparison.

@@

    This simulation is forced by the optimal EPICA palaeo-temperature record
    (Sect.~\ref{sec:paltemp}) and includes palaeo-precipitation reductions
    (Sect.~\ref{sec:palprec}). It uses an horizontal resolution of 1\,km. The
    vertical grid consists of 61~temperature layers in the bedrock and up to
    251~enthalpy layers in the ice, corresponding to vertical resolutions of 50
    and 20\,\unit{m}, respectively.

@@

    The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a,
    blue colour mapping). Fast-flow regions generally occur along the main
    river valleys, while ice domes and ice divides are predominantly located
    over major reliefs (Fig.~\ref{fig:lgmvel}a). Nevertheless, the model
    results depict occasional ice flow across the modern water divides, i.e.
    transfluences.

@@

    These differences in timing of the LGM occur in the model results despite
    the homogeneous temperature and precipitation anomalies supplied as
    palaeo-climate forcing. The LGM timing differences modelled here are
    thus not related to climate, but are inherent to modelled glacier dynamics.
    They result from differences in the subglacial topography, in particular
    catchment sizes and bed roughness, of the different outlet glaciers.

    PISM is a polythermal ice sheet model, i.e. it computes the distribution
    of englacial temperatures resulting from heat advection, vertical
    diffusion in ice and rock, and strain heat release, in three dimensions.
    For large Alpine glaciers flowing in deep valleys, such as the Rhone,
    Rhine and Garda glaciers, several thousand model years are needed to attain
    a thermodynamical equilibrium between cold ice advection from the high
    accumulation areas, upward diffusion of geothermal heat, and heat release
    from strong basal shear strain. Temperature evolution is, in part, limited
    by the slow warming of subglacial bedrock, that has been previously cooled
    downed by subfreezing air temperature before glacier advance.
    For larger glaciers, this thermodynamical equilibrium is typically not yet
    attained around 25\,ka. This is why several Alpine lobes overshoot their
    balanced extent before thinning and receding towards the mountains as they
    warm towards thermodynamical equilibrium.

    Our results indicate that even for a relatively small ice sheet like that
    found in the Alps, the LGM glacier extent corresponds to a transient
    stage, while ka-scale differences in its timing can
    result from complex glacier dynamics. Further differences in LGM timing
    caused by heterogeneous climatic anomalies, such as precipitation
    anomaly gradients, not included in our model set-up, may have
    counterbalanced or emphasize those modelled here.

@@

    In the upper Rhone valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (Fig.~\ref{fig:trimlines}b). This result is
    somewhat depedent on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. However, a previous
    sensitivity study conducted with PISM and a very similar model set-up on
    the Cordillera ice sheet show that varying ice rheological parameters
    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}.

    Beside, our model results depict a polythermal LGM Alpine ice cover.
    Due to sub-freezing temperatures applied in the climate forcing of the
    model, the entire Alpine ice sheet is capped by an upper layer of cold ice.
    In major Alpine valleys, such as the upper Rhone Valley, important ice
    thickness and strain heating contribute to form a layer of temperate ice
    near the glacier base, allowing basal melt, sliding and potentially
    erosion (Fig.~\ref{fig:trimlines}c, white areas). On the other hand, on the
    highest mountains ice cover is to thin and too static to form temperate
    ice, resulting in cold ice down to the bed, preventing potential erosion
    (Fig.~\ref{fig:trimlines}c, hatched areas).

    In the upper Rhone valley, the geographical location of observed trimlines
    show remarkable agreement with the areal extent of regions that are
    modelled to have experienced warm-based for less than 1\,ka (7\,\%) of the
    MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
    detailed comparisons across the
    entire Alpine range and specific sensitivity studies to relevant basal
    sliding and ice rheological parameters, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable. Thus we want to challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.

@@

    \begin{itemize}
      \item The EPICA palaeo-temperature forcing yields maximum ice volume at
            25.2/24.6\,ka (without/with palaeo-precipitation reduction),
            followed by a standstill of major piedmont lobe in the forelands
            until 17.3\,ka, both compatible with much of the dating results,
            whereas the GRIP forcing results in early deglaciation of the
            foreland complete by 21.4\,ka, and the MD01-2444 forcing results in
            a late LGM glaciation peaking at 16.5/15.7\,ka
            (Sect.~\ref{sec:timeseries}).
      \item The EPICA palaeo-temperature forcing yields cumulative ice extent
            compatible with geological evidence during MIS~4 and 2, whereas
            both GRIP and MD01-2444 forcings result in MIS~4 glaciation well
            beyong the mapped LGM ice limits  (Sect.~\ref{sec:footprints}).
    \end{itemize}

@@

    \begin{itemize}
      \item Ice cover is generally underestimated in the Western Alps and
            overestimated in the Eastern Alps, indicating that east-west
            gradients in temperature or precipitation change, absent from our
            model forcing, probably controlled the LGM extent of ice cover in
            the Alps. The asymmetric extent north and south of the Alps
            can be explained by the transient nature of the LGM extent without
            involving north-south gradients in temperature and precipitation
            change (Sect.~\ref{sec:extent}).
      \item The LGM ice flow pattern in the Alps was largely controlled by
            subglacial topography, but several transfluences and two
            self-sustained ice domes may have occurred (Sect.~\ref{sec:flow}).
      \item The LGM was a transient stage in which glaciers were not in
            equilibrium with climate. Timing potentially varied across the
            range due to inherent glacier dynamics (Sect.~\ref{sec:timing}).
      \item Ice thickness during the LGM is modelled to be much higher than in
            reconstructions. In average, modelled surface elevation is 861\,m
            above the Rhone Glacier trimlines, which may instead indicate an
            englacial thermal boundary (Sect.~\ref{sec:thickness}).
      \item Alpine glaciers were potentially very dynamic. They quickly
            responded to climate fluctuation and some potentially advanced many
            times over the foreland during the last glacial cycle
            (Sect.~\ref{sec:glaciations}).
    \end{itemize}

@@

    \begin{figure*}
      \centerline{\includegraphics{alpcyc_lr_timeseries}}
      \caption{%
        \textbf{(a)} Temperature offset time-series from ice core and ocean
        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
        the ice sheet model.
        \textbf{(b)} Modelled total ice volume through the last 120~thousand
        years (ka) expressed in meters of sea level equivalent (m~s.l.e.). Gray
        fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
        and MIS~4 according to a~global compilation of benthic
        \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
      \label{fig:timeseries}
    \end{figure*}

@@

    \begin{figure*}
      \centerline{\includegraphics{alpcyc_hr_profiles}}
      \caption{%
        \textbf{(a, c, e, g)} Approximate glacier flowlines drawn by hand
        rougly following selected glacial valley centerlines.
        \textbf{(b, d, f, h)} Evolution of modelled glacier extent in time,
        bilinearly interpolated along the corresponding profiles, showing
        numerous cycles of advance and retreat over the last glacial cycle
        modulated by subglacial topography and catchment geometry.
        Isolated patches indicate periodic surges from tributary glaciers.}
      \label{fig:profiles}
    \end{figure*}

>>>>

    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences and
    self-sustained ice domes. Modelled maximum ice surface is about 900\,m higher
    than observed trimline elevations, yet our simulation predicts little
    erosion at high elevation due to cold ice conditions. Finally, the Last
    Glacial Maximum advance, often considered synchronous, is here modelled as
    a time-transgressive event, with some glacier lobes reaching their maximum
    as early as 27\,ka, and some as late as 21\,ka.

@@

    From the mid-twentieth
    century, palaeoclimate records extracted from deep sea sediments and
    ice cores have provided a much more detailed picture of the Earth
    environmental history \citep[e.g.,][]{Emiliani.1955,
    Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
    indicating neither one, nor four, but rather tens of glacial and
    interglacial periods. During the last 800000 years (800\,ka), these
    \emph{glacial cycles}
    have succeeded each other with a periodicity about 100\,ka \citep{Hays.etal.1976,
    Augustin.etal.2004}. Nevertheless, this global signal is largely governed
    by the North American and Eurasian ice sheet complexes. It is thus unclear
    whether glacier advances and retreats in the Alps were in pace with global
    sea-level fluctuations.

    Besides, the landform record is typically sparse in time and, most often,
    spatially incomplete. Palaeo-ice sheets did not leave a continuous imprint
    on the landscape, and much of this evidence has been overprinted by
    subsequent glacier re-advances and other geomorphological processes
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2006, Kleman.etal.2010}. Dating
    uncertainties typically increase with age, such that dated reconstructions
    are strongly biased towards earlier glaciations \citep{Heyman.etal.2011}.
    The European Alps have been studied in more detail than any other glacial
    region, but they are no exception to these rules. Although sparse
    geological traces indicate that the last glacial cycle may have comprised
    two or three cycles of glacier growth and decay \citep{Preusser.2004,
    Ivy-Ochs.etal.2008}, most glacial features currently left on the foreland
    present a record of the last major glaciation of the Alps, dating from the
    \emph{Last Glacial Maximum} (LGM).

@@

    These simulations use an horizontal resolution of 2\,km. The vertical grid
    consists of 31~temperature layers in the bedrock and up to 126~enthalpy
    layers in the ice, corresponding to vertical resolutions of 100 and
    40\,\unit{m}, respectively. Simulations ran for \todo{how long?}
    using 36~processors.

@@

    For the three palaeo-temperature records and the two palaeo-precipitation
    parametrizations used, the model yields significant ice volume build-up
    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}b), corresponding to documented
    glaciation periods in the Alps \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
    All simulations also yield important glaciations during MIS~5 and 3, but
    their timing and amplitude varies significantly depending on the climate
    forcing used. All six simulations overestimate the late-glacial re-advance
    during the Younger Dryas
    \citep[\todo{timing}, cf. e.g.,][]{Ivy-Ochs.etal.2009}.
    This is partly because the 2\,km resolution is too coarse to resolve
    Younger Dryas Alpine glaciers, but the large total ice volume
    overestimation resulting from the GRIP temperature forcing
    (Fig.~\ref{fig:timeseries}b, blue curves) certainly indicates that the
    vigorous cooling experienced in Greenland is certainly not representative
    of the Alps.

    Geological data from the best documented Alpine piedmont lobes indicate
    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
    potentially re-advanced to within close reach of this maximum extent, until
    as late as 22 to 17\,ka \citep[Fig.~5]{Wirsig.etal.2016}. This is in very
    good agreement with the result of the EPICA simulations, which yield an
    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
    reductions) followed by a standstill until 17.3\,ka
    (Fig.~\ref{fig:timeseries}b, blue curves). The GRIP palaeo-temperature
    forcing, on the other hand, yields two distinct total ice volume maxima
    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
    a late LGM peak ice volume at 16.5/15.7\,ka several thousand years after
    the dated geological evidence.

@@

    Beside this general pattern, there exist differences in glaciated area
    depending on the palaeo-climate forcing used. The MD01-2444, and to a
    greater extent, the GRIP, palaeo-temperature records, tend to overestimate
    MIS~2 ice cover on all peripheral ranges that surround the Alps
    (Fig.~\ref{fig:footprints}a and c). This is particularly true when
    precipitation corrections are applied (bright colours). In fact, both
    palaeo-temperature records have a higher temperature variability and
    contain brief spells of cold temperatures, which are too short to develop a
    fully grown Alpine ice sheet, but long enough to build up ice cover on a
    smaller scale on these peripheral ranges (Fig.~\ref{fig:timeseries}). On
    the other hand, peripheral glaciation modelled using the EPICA record
    (Fig.~\ref{fig:footprints}b), which has less temperature variability,
    is in relatively good agreement with the geomorphological reconstructions.

    The modelled extent of glaciation during MIS~4 depicts a more pronounced
    sensitivity to the choice of palaeo-climate forcing used. Using the GRIP
    and MD01-2444 palaeo-temperature records, Alpine glaciers are modelled to
    extend well beyond the reach of documented LGM end-moraines and glacial
    erratic terrain (Fig.~\ref{fig:footprints}d and f). Because, in both cases,
    the modelled glaciation extent corresponds to brief cold spells in the
    palaeo-temperature forcing (Fig.~\ref{fig:timeseries}a),
    palaeo-precipitation reductions greatly reduce the excessive modelled ice
    cover (Figs.~\ref{fig:timeseries}b and \ref{fig:footprints}d and f, light
    colours). However, with or without palaeo-precipitation corrections,
    the GRIP and MD01-2444 forcing yield
    modelled ice extents (Fig.~\ref{fig:footprints}d and f) and volumes
    (Fig.~\ref{fig:timeseries}b) higher during MIS~4 than
    during MIS~2. This is in contradiction with geological evidence
    \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
    The EPICA palaeo-temperature forcing, on the other hand, yields a MIS~4
    glaciation that is only slightly less expansive than the MIS~2 glaciation
    (Fig.~\ref{fig:footprints}e). The sensitivity of the glaciated area to
    palaeo-precipitation reductions is mostly limited to the southern terminal
    lobes of central-Alpine glaciers, which are slightly less extensive during
    MIS~2 and 4 (Fig.~\ref{fig:footprints}b and e).

    Based on the above considerations on timing of the LGM during MIS~2, and
    modelled ice
    extent during MIS~4, we select EPICA as our optimal palaeo-temperature
    record for a more detailed and higher-resolution comparison of modelled
    glacier dynamics to available geological evidence
    (Sect.~\ref{sec:results}). As a conservative approach in regard to the
    rapid ice volume fluctuations, we choose to include palaeo-precipitation
    corrections in the following comparison.

@@

    This simulation is forced by the optimal EPICA palaeo-temperature record
    (Sect.~\ref{sec:paltemp}) and includes palaeo-precipitation reductions
    (Sect.~\ref{sec:palprec}). It uses an horizontal resolution of 1\,km. The
    vertical grid consists of 61~temperature layers in the bedrock and up to
    251~enthalpy layers in the ice, corresponding to vertical resolutions of 50
    and 20\,\unit{m}, respectively. This simulation ran for 33~days using
    576~processors.

@@

    The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a,
    blue colour mapping). Ice velocities of several hundred metres per year,
    characterising basal sliding, generally occur along the main river valleys,
    while ice domes and ice divides are predominantly located over major
    reliefs areas, where ice moves only by a few metres per year,
    characterising internal deformation without basal sliding
    (Fig.~\ref{fig:lgmvel}a). Nevertheless, the model results depict exceptions
    to this general pattern as occasional ice flow across the modern water
    divides, i.e. transfluences.

@@

    These differences in timing of the LGM occur in the model results despite
    the homogeneous temperature and precipitation anomalies supplied as
    palaeo-climate forcing. The LGM timing differences modelled here are
    thus not related to climate, but are inherent to modelled glacier dynamics.
    They result from differences in the subglacial topography, in particular
    catchment sizes and hypsometry, of the different outlet glaciers.

    For large Alpine glaciers flowing in deep valleys, such as the Rhone,
    Rhine and Garda glaciers, several thousand model years are needed to attain
    a thermodynamical equilibrium between cold ice advection from the high
    accumulation areas, upward diffusion of geothermal heat, and heat release
    from strong basal shear strain. Temperature evolution is, in part, limited
    by the slow warming of subglacial bedrock, that has been previously cooled
    downed by subfreezing air temperature before glacier advance.
    For larger glaciers, this thermodynamical equilibrium is typically not yet
    attained around 25\,ka. This is why several Alpine lobes overshoot their
    balanced extent before thinning and receding towards the mountains as they
    warm towards thermodynamical equilibrium (supplementary animation).

    Our results indicate that even for a relatively small ice sheet like that
    found in the Alps, the LGM glacier extent corresponds to a transient
    stage, while ka-scale inherent differences in its timing can
    result from complex glacier dynamics. Further differences in LGM timing
    caused by heterogeneous climatic anomalies, such as precipitation
    anomaly gradients, not included in our model set-up, may have
    counterbalanced or emphasize those modelled here.

@@

    In the upper Rhone valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (Fig.~\ref{fig:trimlines}b), wich is significant in
    respect to the modelled regional ice thickness. This result is
    somewhat dependent on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. However, a previous
    sensitivity study conducted with PISM and a very similar model set-up on
    the Cordillera ice sheet show that varying ice rheological parameters
    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}.

    Beside, our model results depict a polythermal LGM Alpine ice cover.
    Due to sub-freezing temperatures applied in the climate forcing of the
    model, the entire Alpine ice sheet is capped by an upper layer of cold ice.
    In major Alpine valleys, such as the upper Rhone Valley, important ice
    thickness and strain heating contribute to form a layer of temperate ice
    near the glacier base, allowing basal melt, sliding and potentially
    erosion (Fig.~\ref{fig:trimlines}c, white areas). On the other hand, on the
    highest mountains ice cover is too thin and too static to form temperate
    ice, resulting in cold ice down to the bed, preventing potential erosion
    (Fig.~\ref{fig:trimlines}c, hatched areas).

    In the upper Rhone valley, the geographical location of observed trimlines
    show remarkable agreement with the areal extent of regions that are
    modelled to have experienced warm-based for less than 1\,ka (7\,\%) of the
    MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
    detailed comparisons across the
    entire Alpine range and specific sensitivity studies to relevant basal
    sliding and ice rheological parameters, the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable. In this context, the model results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.

@@

    \begin{itemize}
      \item The EPICA palaeo-temperature forcing yields maximum ice volume at
            25.2/24.6\,ka (without/with palaeo-precipitation reduction),
            followed by a standstill of major piedmont lobe in the forelands
            until 17.3\,ka, both compatible with much of the dating results,
            whereas the GRIP forcing results in early deglaciation of the
            foreland complete by 21.4\,ka, and the MD01-2444 forcing results in
            a late LGM glaciation peaking at 16.5/15.7\,ka
            (Sect.~\ref{sec:timeseries}).
      \item The EPICA palaeo-temperature forcing yields cumulative ice extent
            compatible with geological evidence during MIS~4 and 2, whereas
            both GRIP and MD01-2444 forcings result in MIS~4 glaciation well
            beyond the mapped LGM ice limits  (Sect.~\ref{sec:footprints}).
    \end{itemize}

@@

    \begin{itemize}
      \item Ice cover is generally underestimated in the Western Alps and
            overestimated in the Eastern Alps, indicating that east-west
            gradients in temperature or precipitation change, absent from our
            model forcing, probably controlled the LGM extent of ice cover in
            the Alps. The asymmetric extent north and south of the Alps
            can be explained by the transient nature of the LGM extent without
            involving north-south gradients in temperature and precipitation
            change (Sect.~\ref{sec:extent}).
      \item The LGM ice flow pattern in the Alps was largely controlled by
            subglacial topography, but several transfluences and two
            self-sustained ice domes may have occurred (Sect.~\ref{sec:flow}).
      \item The LGM was a transient stage in which glaciers were not in
            equilibrium with climate. Its timing potentially varied across the
            range due to inherent glacier dynamics (Sect.~\ref{sec:timing}).
      \item Ice thickness during the LGM is modelled to be much higher than in
            reconstructions. In average, modelled surface elevation is 861\,m
            above the Rhone Glacier trimlines, which may instead indicate an
            englacial thermal boundary (Sect.~\ref{sec:thickness}).
      \item Alpine glaciers were potentially very dynamic. They quickly
            responded to climate fluctuation and some potentially advanced many
            times over the foreland during the last glacial cycle
            (Sect.~\ref{sec:glaciations}).
    \end{itemize}

@@

    \begin{figure*}
      \centerline{\includegraphics{alpcyc_lr_timeseries}}
      \caption{%
        \textbf{(a)} Temperature offset time-series from ice core and ocean
        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
        the ice sheet model.
        \textbf{(b)} Modelled total ice volume through the last 120~thousand
        years (ka) expressed in meters of sea level equivalent (m~s.l.e.). Gray
        fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
        and MIS~4 according to a~global compilation of benthic
        \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. The simulation
        driven by the EPICA temperature yields smaller ice volume variability
        and a more realistic timing of deglaciation.}
      \label{fig:timeseries}
    \end{figure*}

@@

    \begin{figure*}
      \centerline{\includegraphics{alpcyc_hr_profiles}}
      \caption{%
        \textbf{(a, c, e, g)} Approximate glacier flowlines drawn by hand
        rougly following selected glacial valley centerlines for the Rhine,
        Rhone, Ivrea and Isère glaciers.
        \textbf{(b, d, f, h)} Evolution of modelled glacier extent in time,
        bilinearly interpolated along the corresponding profiles, showing
        numerous cycles of advance and retreat over the last glacial cycle
        modulated by subglacial topography and catchment geometry.
        Isolated patches indicate periodic surges from tributary glaciers.}
      \label{fig:profiles}
    \end{figure*}
